A Linear Inside-Outside Algorithm for Correcting Sequencing Errors in Structured RNA

### 2013-01-09 Recomb/introduction_RECOMB.tex
shrunk figures ... could do better but got to go

For half a century, biological molecules have been studied as a proxy to understand evolution~\cite{Zuckerkandl1965}, and due
to their fundamental functions and remarkably conserved structures, rRNAs have always been a prime candidate for phylogenetic
studies~\cite{Olsen1986, Olsen1993}. In recent years, studies as the \emph{Human Microbiome Project}~\cite{Turnbaugh2007} benefited
of new technologies such as the NGS techniques to sequence as many new organisms as possible and extract an unprecedented flow of new information. 
Nonetheless, these high-throughput techniques typically have high error rates that make their applications to metagenomics (a.k.a. environmental
genomics) studies challenging. For instance, pyrosequencing as implemented by Roche's 454 produces may have an error rate raising up to 10\%.
Because there is no cloning step, resequencing to increase accuracy is not possible and it is therefore vital to disentangle noise from true sequence
diversity in this type of data \cite{Quince:2009uq}. Errors can be significantly reduced  when large multiple sequence alignments with close homologs
are available, but in studies of new or not well known organisms, such information is rather sparse. In particular, it is common that there is not enough  similarity to 
differentiate between the sequencing errors and the natural polymorphisms that we want to observe, often inflating the diversity estimates~\cite{Kunin2010}.
A few techniques have been developed to remedy to this problem~\cite{Quinlan2008,Medvedev2011} but they do not take
into account all the available information. It is therefore essential to develop methods that can exploit any type of signal available to correct errors.

>>>>

For half a century, biological molecules have been studied as a proxy to understand evolution~\cite{Zuckerkandl1965}, and due
to their fundamental functions and remarkably conserved structures, rRNAs have always been a prime candidate for phylogenetic
studies~\cite{Olsen1986, Olsen1993}. In recent years, studies as the \emph{Human Microbiome Project}~\cite{Turnbaugh2007} benefited
of new technologies such as the NGS techniques to sequence as many new organisms as possible and extract an unprecedented flow of new information. 
Nonetheless, these high-throughput techniques typically have high error rates that make their applications to metagenomics (a.k.a. environmental
genomics) studies challenging. For instance, pyrosequencing as implemented by Roche's 454 produces may have an error rate raising up to 10\%.
Because there is no cloning step, resequencing to increase accuracy is not possible and it is therefore vital to disentangle noise from true sequence
diversity in this type of data \cite{Quince:2009uq}. Errors can be significantly reduced  when large multiple sequence alignments with close homologs
are available, but in studies of new or not well known organisms, such information is rather sparse. In particular, it is common that there is not enough  similarity to 
differentiate between the sequencing errors and the natural polymorphisms that we want to observe, often leading to artificially inflated diversity estimates~\cite{Kunin2010}.
A few techniques have been developed to remedy to this problem~\cite{Quinlan2008,Medvedev2011} but they do not take
into account all the available information. It is therefore essential to develop methods that can exploit any type of signal available to correct errors.

### 2013-01-09 Recomb/methods_RECOMB.tex
shrinking equations
shrunk figures ... could do better but got to go

\subsubsection{Energy Contribution}
The free-energy contribution in our pseudo-energy model corresponds to an additive stacking-pairs model, taking values from the Turner 2004 model retrieved from the NNDB~\cite{Turner2010}. Given a candidate sequence $s$ for a secondary structure $S$, the free-energy of $S$ on $s$ is given by
\begin{align*}
  \ES(s,S) = \sum_{\substack{(i,j)\to (i',j')\in S\\ \text{stacking pairs}}}\ES^{\beta}_{s_is_j\to s_{i'}s_{j'}} 
\end{align*}
where $\ES^{\beta}_{ab\to a'b'}$ is set to $0$ if $ab=\varnothing$ (no base-pair to stack onto), the tabulated free-energy of stacking pairs $(ab)/(a'b')$ in the Turner model if available, or $\beta\in[0,\infty]$ for non-Watson-Crick/Wobble entries (i.e. not in $\{\Gb\Ub,\Ub\Gb,\Cb\Gb,\Gb\Cb, \Ab\Ub\text{ or }\Ub\Ab\}$). This latter parameter allows to choose whether to simply penalize invalid base pairs, or forbid them altogether ($\beta = \infty$).
The loss of precision due to this simplification of the Turner model remains reasonable since the targeted secondary structure is fixed 
(e.g. multiloops do not account for base-specific contributions). Furthermore, it greatly eases the design and implementation of dynamic-programming equations. 
\subsubsection{Isostericity Contribution}
The concept of isostericity score~\cite{Stombaugh2009} is based on the geometric discrepancy (superimposability) of two base-pairs, using individual additive contributions computed by Stombaugh~\emph{et al}~\cite{Stombaugh2009}. Let $s$ be a candidate sequence for a secondary structure $S$, given in the context of a gap-free RNA alignment $\Omega$,  we define the contribution of the isostericity to the pseudo-energy as
\begin{align*}
  \ES(s,S,\Omega) &= \sum_{\substack{(i,j)\in S\\ \text{pairs}}}\EI^{\Omega}_{(i,j),s_i s_j}, & \text{where}&& 	\EI^{\Omega}_{(i,j),ab}:=
	\frac{
		\sum_{s'\in\Omega}
			\text{\ISO}((s_i',s_j'),(a,b))}
%-		\left(			\text{\ISO}((s_i',s_j'),(s_i,s_j))				\right)	
{		
		|\Omega|
	}
\end{align*}
is the average isostericity of a base-pair in the candidate sequence, compared with the reference alignment.
The $\ISO$ function uses the {Watson-Crick/Watson-Crick} cis isostericity matrix computed by Stombaugh~\emph{et al}~\cite{Stombaugh2009}. Isostericity scores range between $0$ and $9.7$, with $0$ being assigned to a perfect isostericity, and a penalty of $10$ is used for missing entries.
The isostericity contribution will favor exponentially sequences that are likely to adopt a similar local conformation as the sequences contained in the alignment.

@@

\subsubsection{Definitions}
Let $B:=\left\{\Ab,\Cb,\Gb,\Ub\right\}$ be the set of nucleotides.
Given $s\in B^n$ an RNA sequence, let $s_i$ be the nucleotide at position $i$. Let $\Omega$ be a set of un-gapped RNA sequences of
length $n$, and $S$ a secondary structure without pseudoknots. 
Formally, if $(i,j)$ and $(k,l)$ are base pairs in $S$, there is no overlapping extremities
 $\{i,j\}\cap \{k,l\}=\varnothing$ and either the intersection is empty 
 ($[i,j]\cap[k,l]=\varnothing$) or one is included in the other ($[k,l]\subset[i,j]$ or 
 $[i,j]\subset[k,l])$.

@@

\subsubsection{Inside}
\begin{figure}
\resizebox{\textwidth}{!}{\input{FigDPInside.pgf}}
\caption{Principle of the inside computation (aka partition function). Any sequence with $m$ mutations over an interval $[i,j]$ 
can be decomposed as a sequence over $[i+1,j]$ preceded by a, possibly mutated, base at $i$ 
(Unpaired case), a sequence over $[i+1,j-1]$ surrounded by some base-pair (Stacking-pair case), 
or as two sequences over $[i+1,k-1]$ and $[k+1,j]$, completed by some base-pair (General base-pairing case). 
In each case, one has to investigate any possible ways to distribute mutations between the different sequences 
and locally instanciated bases.}
\end{figure}

@@

The main recursion is composed of four terms:
\begin{equation}
	\Z{i,j}{m}{a,b}:=\left\{
  \begin{array}{ll}
  		\displaystyle
      \sum_{\substack{a'\in \B,\\ \Kron_{a',s_i}\le m}}  
      \Z{i+1,j}{m-\Kron_{a',s_i}}{a',b} & \text{If }S_{i}=-1\\
      \displaystyle
      \sum_{\substack{a',b'\in \B^2,\\ \Kron_{a'b',s_is_j}\le m}}
			 e^{\frac{-E_{(i,j),ab \to a'b'}^{\Omega,\beta}}{RT}}
			 \cdot \Z{i+1,j-1}{m-\Kron_{a'b',s_is_j}}{a',b'}&
			 \text{Elif }S_i=j \land S_{i-1}=j+1\\
			 \displaystyle
      \sum_{\substack{a',b'\in \B^2,\\ \Kron_{a'b',s_is_k}\le m}}
      \sum_{m'=0}^{m-\Kron_{a'b',s_is_k}}
   		 e^{\frac{-E_{(i,k),\varnothing\to a'b'}^{\Omega,\beta}}{RT}}
      \cdot\Z{i+1,k-1}{m-\Kron_{a'b',s_is_k}-m'}{a',b'}
      \cdot\Z{k+1,j}{m'}{b',b} & \text{Elif }S_i=k \land i < k \leq j\\
      0 &\text{Otherwise}
	\end{array}\right.
\label{eq:Z_rec}
\end{equation}
The cases can be broken down as follows:
\begin{description}
\item[$S_{i}=-1$:] If the nucleotide at position $i$ is unpaired, then 
any sequence consists in a, possibly mutated, nucleotide $a'$ at position  $i$, 
followed by a sequence over $[i+1,j]$ having either $m-\Kron_{a',s_i}$, accounting for a possible mutation 
at position $i$, and having flanking nucleotides $a'$ and $b$.
\item[$S_i=j$ and $S_{i-1}=j+1$:] Any sequence generated in $[i,j]$ consists of two, possibly mutated, nucleotides $a'$ and $b'$, flanking a sequence over $[i+1,j-1]$ having distance $m-\Kron_{a'b',s_is_j}$ (to avoid exceeding the targeted distance $m$).
Since positions $i$ and $i-1$ are paired with $j$ and $j+1$ respectively, 
then a stacking energy contribution is added. 
\item[$S_i=k$ and $i<k \leq j$:] If position $i$ is paired and not involved in a stacking, then the 
only term contributing directly to the energy is the isostericity of the base pair $(i,k)$. 
Any sequence on $[i,j]$ consists of two nucleotide $a'$ and $b'$ at positions $i$ and $k$ respectively, flanking a sequence over interval $[i+1,k-1]$ and preceding a (possibly empty) sequence interval $[k+1,j]$. Since the number of mutations sum to $m$ over the whole sequence must , then a parameter $m'$ is introduced to distribute the remaining mutations between the two sequences.
\item[Else:] In any other case, we are in a derivation of the SCFG that does not correspond to the secondary structure $S$, and we return $0$.
\end{description}

\subsubsection{Outside}	
\begin{figure}
\resizebox{\textwidth}{!}{\input{FigDPOutside.pgf}}
\caption{Principle of the outside computation. Note that the outside algorithm uses intermediate results from the inside algorithm, 
therefore its efficient implementation requires an implementation of the inside computation.}
\end{figure}
The \emph{Outside} function, $\mathcal Y$, is the partition function considering only the 
contributions of subsequences $[0,i]\cup[j,n-1]$ over the mutants of $s$ having exactly $m$ mutations between $[0,i]\cup[j,n-1]$ and whose nucleotide at position $i+1$ is $a$ 
(resp. in position $j-1$ it is $b$).
We define function $\Y{i,j}{m}{a,b}$ as a recurrence, and will use as initial conditions:
\begin{equation}
	\Y{-1,j}{m}{X,X}:=
		\displaystyle
	  \Z{j,n-1}{m}{X,X}
\label{eq:Y_in}
\end{equation}
The recurrence, as shown below, will increase the interval $[i,j]$ by decreasing $i$ when
it is not base paired. If it is with a position $k>j$, we increase $j$ to include it.
 Thus, when we need
to evaluate an interval as $(-1,j)$, all stems between $(0,j)$ are taken into account and the
structure between $(j,n-1)$ must be a set of independent stems. Therefore,
 all the outside energy between $[j,n-1]$ is
equal to $\Z{j,n-1}{m}{X,X}$, for any $X\in B$. The recursion itself is as follows.
\begin{equation}
	\Y{i,j}{m}{a,b} = \left\{
  \begin{array}{ll}
		\displaystyle
    \sum_{\substack{a'\in \B,\\ \Kron_{a',s_i}\le m}}
    \Y{i-1,j}{m- \Kron_{a',s_i}}{a',b} &
    \text{Elif }S_i=-1 \\
    \displaystyle
    \sum_{\substack{a'b'\in \B^2,\\ \Kron_{a'b',s_is_j}\le m}}
		 e^{\frac{-E_{(i,j),ab \to a'b'}^{\Omega,\beta}}{RT}}\cdot
    \Y{i-1,j+1}{m- \Kron_{a'b',s_is_j}}{a',b'} &
   	 \text{Elif }S_{i}=j \land S_{i+1}=j-1\\
		 \displaystyle
		 \sum_{\substack{a'b'\in \B^2,\\ \Kron_{a'b',s_is_k}\le m}}
		 \sum_{m'=0}^{m-\Kron_{a'b',s_is_k}}
  		 e^{\frac{-E_{(i,k),\varnothing\to a'b'}^{\Omega,\beta}}{RT}}
		 \cdot\Y{i-1,k+1}{m- \Kron_{a'b',s_is_k} - m'}{a',b'}
     \cdot\Z{j,k-1}{m'}{b,b'} &
		 \text{Elif }S_{i}=k \geq j\\
		 \displaystyle
		 \sum_{\substack{a'b'\in \B^2,\\ \Kron_{a'b',s_ks_i}\le m}}
		 \sum_{m'=0}^{m-\Kron_{a'b',s_ks_i}}
   	 e^{\frac{-E_{(k,i),\varnothing\to a'b'}^{\Omega,\beta}}{RT}}
		 \cdot\Y{k-1,j}{m- \Kron_{a'b',s_ks_i} - m'}{a',b}
     \cdot\Z{k+1,i-1}{m'}{a',b'} &
		 \text{Elif }-1 < S_{i}=k < i\\
		 0 & \text{Otherwise}
  \end{array}\right.
\label{eq:Y_rec}
\end{equation}
The five cases can be broked down as follows.
\begin{description}
\item[$S_i=-1$:] If the nucleotide at position $i$ is not paired, then the value is the same
as if we decrease the lower interval bound by $1$ (i.e. $i-1$), and consider all possible
nucleotides $a'$ at position $i$, correcting the number of mutants
in function of $\Kron_{a',s_i}$.
\item[$S_{i}=j$ and $S_{i+1}=j-1$:] If nucleotide $i$ is paired with $j$ and nucleotide $i+1$ is
paired with $j-11$, we are in the only case were stacked base pairs can occur. We thus add
the energy of the stacking and of the isostericity of the base pair $(i,j)$. What is left
to compute is the \emph{outside} value for the interval $[i-1,j+1]$ over all possible nucleotides 
$a',b'\in B^2$ at positions $i$ and $j$ respectively.
\item[$S_{i}=k \geq j$:]If nucleotide $i$ is paired with position $k\geq j$, 
and is not stacked inside, the 
only term contributing directly to the energy is the isostericity of the base pair $(i,k)$. 
We can then consider the outside interval $[i-1,k+1]$ by multiplying it by the the \emph{inside}
value of the newly included interval (i.e. $[j,k-1]$), for 
all possible values $a',b'\in B^2$ for nucleotides at positions $i$ and $k$ respectively.
\item[$-1<S_{i}<i$:]As above but if position $i$ is to pairing with a lower value.
\item[Else:] In all other cases, we are in a derivation of the SCFG that does not correspond to the 
secondary structure $S$, and we return $0$.

@@

\subsubsection{Combining Inside and Outside Values into Point-Wise Mutations Probabilities}
By construction, the partition function over all sequences at exactly $m$ mutations of $s$ can 
be described in function of the \emph{inside} term as $\Z{0,n-1}{m}{X,X}$,
 for any nucleotide $X\in B$ or
in function of the \emph{outside} term, for any unpaired position $k$:
$$
	\Z{0,n-1}{m}{X,X}
	\equiv
	\sum_{\substack{a\in \B,\\ \Kron_{a,s[k]}\le m}}	
	\Y{k-1,k+1}{m-\Kron_{a,s[k]}}{a,a}
$$

We are now left to compute the probability that a given position is a given nucleotide.
We leverage the \emph{Inside-Outside} construction to immediately obtain the following $3$ cases.
Given $i\in[0,n-1],x\in B$, and $M\geq 0$ a bound on the number of allowed mutations. 
\begin{align}
	\mathbb{P}(s_i = x\mid s,\Omega, S,M) &:= \frac{\mathcal{W}(i,x,s,\Omega,S,M)}{\sum_{m=0}^{M}\Z{0,n-1}{m}{X,X}}\label{eq:normalize}\\ 
\mathcal{W}(i,x,s,\Omega,S,M)&=
 \left\{
	\begin{array}{ll}
			\sum_{m=0}^{M}
			\Y{i-1,i+1}{m-\Kron_{x,s_i}}{x,x}
		&\text{If }S_i = -1\\
			\displaystyle
			\sum_{m=0}^{M}
			\sum_{\substack{b\in B\\\Kron_{xb,s_is_k\leq m}}}
			\sum_{m'=0}^{m-\Kron_{xb,s_is_k}}
     	 e^{\frac{-E_{(i,k),\varnothing\to xb}^{\Omega,\beta} }{RT}}
			\cdot\Y{i-1,k+1}{m-\Kron_{xb,s_is_k-m'}}{x,b}
			\cdot\Z{i+1,k-1}{m'}{x,b}
		&\text{If }S_i=k>i\\
    \displaystyle
			\sum_{m=0}^{M}
			\sum_{\substack{b\in B\\\Kron_{bx,s_ks_i\leq m}}}
			\sum_{m'=0}^{m-\Kron_{bx,s_ks_i}}
     	 e^{\frac{-E^{\Omega, \beta}_{(k,i),\varnothing\to bx}}{RT}}
			\cdot\Y{k-1,i+1}{m-\Kron_{bx,s_ks_i-m'}}{b,x}
			\cdot\Z{k+1,i-1}{m'}{b,x}
		&\text{If }S_i=k<i
	\end{array}\right.\label{eq:combine}
\end{align}

>>>>

\subsubsection{Energy Contribution.}
The free-energy contribution in our pseudo-energy model corresponds to an additive stacking-pairs model, taking values from the Turner 2004 model retrieved from the NNDB~\cite{Turner2010}. Given a candidate sequence $s$ for a secondary structure $S$, the free-energy of $S$ on $s$ is given by
\begin{align*}
  \ES(s,S) = \sum_{\substack{(i,j)\to (i',j')\in S\\ \text{stacking pairs}}}\ES^{\beta}_{s_is_j\to s_{i'}s_{j'}} 
\end{align*}
where $\ES^{\beta}_{ab\to a'b'}$ is set to $0$ if $ab=\varnothing$ (no base-pair to stack onto), the tabulated free-energy of stacking pairs $(ab)/(a'b')$ in the Turner model if available, or $\beta\in[0,\infty]$ for non-Watson-Crick/Wobble entries (i.e. neither $\Gb\Ub$, $\Ub\Gb$, $\Cb\Gb$, $\Gb\Cb$, $\Ab\Ub$ nor $\Ub\Ab$). This latter parameter allows to choose whether to simply penalize invalid base pairs, or forbid them altogether ($\beta = \infty$).
The loss of precision due to this simplification of the Turner model remains reasonable since the targeted secondary structure is fixed 
(e.g. multiloops do not account for base-specific contributions). Furthermore, it greatly eases the design and implementation of dynamic-programming equations. 
\subsubsection{Isostericity Contribution.}
The concept of isostericity score~\cite{Stombaugh2009} is based on the geometric discrepancy (superimposability) of two base-pairs, using individual additive contributions computed by Stombaugh~\emph{et al}~\cite{Stombaugh2009}. Let $s$ be a candidate sequence for a secondary structure $S$, given in the context of a gap-free RNA alignment $\Omega$,  we define the contribution of the isostericity to the pseudo-energy as
\begin{align*}
  \ES(s,S,\Omega) &= \sum_{\substack{(i,j)\in S\\ \text{pairs}}}\EI^{\Omega}_{(i,j),s_i s_j}, & \text{where}&& 	\EI^{\Omega}_{(i,j),ab}:=
	\frac{
		\sum_{s'\in\Omega}
			\text{\ISO}((s_i',s_j'),(a,b))}
%-		\left(			\text{\ISO}((s_i',s_j'),(s_i,s_j))				\right)	
{		
		|\Omega|
	}
\end{align*}
is the average isostericity of a base-pair in the candidate sequence, compared with the reference alignment.
The $\ISO$ function uses the {Watson-Crick/Watson-Crick} cis isostericity matrix computed by Stombaugh~\emph{et al}~\cite{Stombaugh2009}. Isostericity scores range between $0$ and $9.7$, $0$ corresponding to a perfect isostericity, and a penalty of $10$ is used for missing entries.
The isostericity contribution will favor exponentially sequences that are likely to adopt a similar local conformation as the sequences contained in the alignment.

@@

\subsubsection{Definitions.}
Let $B:=\left\{\Ab,\Cb,\Gb,\Ub\right\}$ be the set of nucleotides.
Given $s\in B^n$ an RNA sequence, let $s_i$ be the nucleotide at position $i$. Let $\Omega$ be a set of un-gapped RNA sequences of
length $n$, and $S$ a secondary structure without pseudoknots. 
Formally, if $(i,j)$ and $(k,l)$ are base pairs in $S$, there is no overlapping extremities
 $\{i,j\}\cap \{k,l\}=\varnothing$ and either the intersection is empty 
 ($[i,j]\cap[k,l]=\varnothing$) or one is included in the other ($[k,l]\subset[i,j]$ or 
 $[i,j]\subset[k,l])$.

@@

\subsubsection{Inside computation.}
\begin{figure}[t]
\begin{minipage}{0.60\textwidth}
\resizebox{\textwidth}{!}{\input{FigDPInside.pgf}}
\end{minipage}
\begin{minipage}{0.38\textwidth}
\caption{Principle of the inside computation (aka partition function). Any sequence with $m$ mutations over an interval $[i,j]$ 
can be decomposed as a sequence preceded by a, possibly mutated, base at $i$ 
(Unpaired case), a sequence surrounded by some base-pair (Stacking-pair case), 
or as two sequences separated by some base-pair (General base-pairing case). 
In each case, one has to investigate any possible ways to distribute mutations between the different sequences 
and locally instantiated bases.}
\end{minipage}
\end{figure}

@@

The main recursion is composed of four terms:\\

\begin{equation}
\label{eq:Z_rec}
\resizebox{\textwidth}{!}{%
	$\Z{i,j}{m}{a,b}:=\left\{
  \begin{array}{ll}
  		\displaystyle
      \sum_{\substack{a'\in \B,\\ \Kron_{a',s_i}\le m}}  
      \Z{i+1,j}{m-\Kron_{a',s_i}}{a',b} & \text{If }S_{i}=-1\\
      \displaystyle
      \sum_{\substack{a',b'\in \B^2,\\ \Kron_{a'b',s_is_j}\le m}}
			 e^{\frac{-E_{(i,j),ab \to a'b'}^{\Omega,\beta}}{RT}}
			 \cdot \Z{i+1,j-1}{m-\Kron_{a'b',s_is_j}}{a',b'}&
			 \text{Elif }S_i=j \land S_{i-1}=j+1\\
			 \displaystyle
      \sum_{\substack{a',b'\in \B^2,\\ \Kron_{a'b',s_is_k}\le m}}
      \sum_{m'=0}^{m-\Kron_{a'b',s_is_k}}
   		 e^{\frac{-E_{(i,k),\varnothing\to a'b'}^{\Omega,\beta}}{RT}}
      \cdot\Z{i+1,k-1}{m-\Kron_{a'b',s_is_k}-m'}{a',b'}
      \cdot\Z{k+1,j}{m'}{b',b} & \text{Elif }S_i=k \land i < k \leq j\\
      0 &\text{Otherwise}
	\end{array}\right.$}
\end{equation}
\
The cases can be broken down as follows:
\begin{description}
\item[$S_{i}=-1$:] If the nucleotide at position $i$ is unpaired, then 
any sequence consists in a, possibly mutated, nucleotide $a'$ at position  $i$, 
followed by a sequence over $[i+1,j]$ having either $m-\Kron_{a',s_i}$, accounting for a possible mutation 
at position $i$, and having flanking nucleotides $a'$ and $b$.
\item[$S_i=j$ and $S_{i-1}=j+1$:] Any sequence generated in $[i,j]$ consists of two, possibly mutated, nucleotides $a'$ and $b'$, flanking a sequence over $[i+1,j-1]$ having distance $m-\Kron_{a'b',s_is_j}$ (to avoid exceeding the targeted distance $m$).
Since positions $i$ and $i-1$ are paired with $j$ and $j+1$ respectively, 
then a stacking energy contribution is added. 
\item[$S_i=k$ and $i<k \leq j$:] If position $i$ is paired and not involved in a stacking, then the 
only term contributing directly to the energy is the isostericity of the base pair $(i,k)$. 
Any sequence on $[i,j]$ consists of two nucleotide $a'$ and $b'$ at positions $i$ and $k$ respectively, flanking a sequence over interval $[i+1,k-1]$ and preceding a (possibly empty) sequence interval $[k+1,j]$. Since the number of mutations sum to $m$ over the whole sequence must , then a parameter $m'$ is introduced to distribute the remaining mutations between the two sequences.
\item[Else:] In any other case, we are in a derivation of the SCFG that does not correspond to the secondary structure $S$, and we return $0$.
\end{description}

\subsubsection{Outside computation.}	
\begin{figure}[t]
\begin{minipage}{0.75\textwidth}
\resizebox{\textwidth}{!}{\input{FigDPOutside.pgf}}
\end{minipage}
\begin{minipage}{0.24\textwidth}
\caption{Principle of the outside computation. Note that the outside algorithm uses intermediate results from the inside algorithm, 
therefore its efficient implementation requires a precomputation of the inside contributions.}
\end{minipage}
\end{figure}
The \emph{Outside} function, $\mathcal Y$, is the partition function considering only the 
contributions of subsequences $[0,i]\cup[j,n-1]$ over the mutants of $s$ having exactly $m$ mutations between $[0,i]\cup[j,n-1]$ and whose nucleotide at position $i+1$ is $a$ 
(resp. in position $j-1$ it is $b$).
We define function $\Y{i,j}{m}{a,b}$ as a recurrence, and will use as initial conditions:
\begin{equation}
	\Y{-1,j}{m}{X,X}:=
		\displaystyle
	  \Z{j,n-1}{m}{X,X}
\label{eq:Y_in}
\end{equation}
The recurrence, as shown below, will increase the interval $[i,j]$ by decreasing $i$ when
it is not base paired. If it is with a position $k>j$, we increase $j$ to include it.
 Thus, when we need
to evaluate an interval as $(-1,j)$, all stems between $(0,j)$ are taken into account and the
structure between $(j,n-1)$ must be a set of independent stems. Therefore,
 all the outside energy between $[j,n-1]$ is
equal to $\Z{j,n-1}{m}{X,X}$, for any $X\in B$. The recursion itself unfolds as follows.
\begin{equation}
\resizebox{\textwidth}{!}{%
	$\Y{i,j}{m}{a,b} = \left\{
  \begin{array}{ll}
		\displaystyle
    \sum_{\substack{a'\in \B,\\ \Kron_{a',s_i}\le m}}
    \Y{i-1,j}{m- \Kron_{a',s_i}}{a',b} &
    \text{Elif }S_i=-1 \\
    \displaystyle
    \sum_{\substack{a'b'\in \B^2,\\ \Kron_{a'b',s_is_j}\le m}}
		 e^{\frac{-E_{(i,j),ab \to a'b'}^{\Omega,\beta}}{RT}}\cdot
    \Y{i-1,j+1}{m- \Kron_{a'b',s_is_j}}{a',b'} &
   	 \text{Elif }S_{i}=j \land S_{i+1}=j-1\\
		 \displaystyle
		 \sum_{\substack{a'b'\in \B^2,\\ \Kron_{a'b',s_is_k}\le m}}
		 \sum_{m'=0}^{m-\Kron_{a'b',s_is_k}}
  		 e^{\frac{-E_{(i,k),\varnothing\to a'b'}^{\Omega,\beta}}{RT}}
		 \cdot\Y{i-1,k+1}{m- \Kron_{a'b',s_is_k} - m'}{a',b'}
     \cdot\Z{j,k-1}{m'}{b,b'} &
		 \text{Elif }S_{i}=k \geq j\\
		 \displaystyle
		 \sum_{\substack{a'b'\in \B^2,\\ \Kron_{a'b',s_ks_i}\le m}}
		 \sum_{m'=0}^{m-\Kron_{a'b',s_ks_i}}
   	 e^{\frac{-E_{(k,i),\varnothing\to a'b'}^{\Omega,\beta}}{RT}}
		 \cdot\Y{k-1,j}{m- \Kron_{a'b',s_ks_i} - m'}{a',b}
     \cdot\Z{k+1,i-1}{m'}{a',b'} &
		 \text{Elif }-1 < S_{i}=k < i\\
		 0 & \text{Otherwise}
  \end{array}\right.$
}\label{eq:Y_rec}
\end{equation}
The five cases can be broked down as follows.
\begin{description}
\item[$S_i=-1$:] If the nucleotide at position $i$ is not paired, then the value is the same
as if we decrease the lower interval bound by $1$ (i.e. $i-1$), and consider all possible
nucleotides $a'$ at position $i$, correcting the number of mutants
in function of $\Kron_{a',s_i}$.
\item[$S_{i}=j$ and $S_{i+1}=j-1$:] If nucleotide $i$ is paired with $j$ and nucleotide $i+1$ is
paired with $j-11$, we are in the only case were stacked base pairs can occur. We thus add
the energy of the stacking and of the isostericity of the base pair $(i,j)$. What is left
to compute is the \emph{outside} value for the interval $[i-1,j+1]$ over all possible nucleotides 
$a',b'\in B^2$ at positions $i$ and $j$ respectively.
\item[$S_{i}=k \geq j$:]If nucleotide $i$ is paired with position $k\geq j$, 
and is not stacked inside, the 
only term contributing directly to the energy is the isostericity of the base pair $(i,k)$. 
We can then consider the outside interval $[i-1,k+1]$ by multiplying it by the the \emph{inside}
value of the newly included interval (i.e. $[j,k-1]$), for 
all possible values $a',b'\in B^2$ for nucleotides at positions $i$ and $k$ respectively.
\item[$-1<S_{i}<i$:]As above but if position $i$ is to pairing with a lower value.
\item[Else:] In all other cases, we are in a derivation of the SCFG that does not correspond to the 
secondary structure $S$, and we return $0$.

@@

\subsubsection{Combining Inside and Outside Values into Point-Wise Mutations Probabilities.}
By construction, the partition function over all sequences at exactly $m$ mutations of $s$ can 
be described in function of the \emph{inside} term as $\Z{0,n-1}{m}{X,X}$,
 for any nucleotide $X\in B$ or
in function of the \emph{outside} term, for any unpaired position $k$:
$$
	\Z{0,n-1}{m}{X,X}
	\equiv
	\sum_{\substack{a\in \B,\\ \Kron_{a,s[k]}\le m}}	
	\Y{k-1,k+1}{m-\Kron_{a,s[k]}}{a,a}
$$

We are now left to compute the probability that a given position is a given nucleotide.
We leverage the \emph{Inside-Outside} construction to immediately obtain the following $3$ cases.
Given $i\in[0,n-1],x\in B$, and $M\geq 0$ a bound on the number of allowed mutations. 
\begin{equation}
	\mathbb{P}(s_i = x\;|\; M) := \frac{\mathcal{W}^M_{\substack{i, [x]}}}{\sum_{m=0}^{M}\Z{0,n-1}{m}{X,X}}\label{eq:normalize}
\end{equation}
where $\mathcal{W}{\substack{\ast\\ \ast}}$ is defined by:
\begin{equation}
\resizebox{\textwidth}{!}{%
$ \mathcal{W}^M_{\substack{i, [x]}} =  \left\{
	\begin{array}{ll}
			\sum_{m=0}^{M}
			\Y{i-1,i+1}{m-\Kron_{x,s_i}}{x,x}
		&\text{If }S_i = -1\\
			\displaystyle
			\sum_{m=0}^{M}
			\sum_{\substack{b\in B\\\Kron_{xb,s_is_k\leq m}}}
			\sum_{m'=0}^{m-\Kron_{xb,s_is_k}}
     	 e^{\frac{-E_{(i,k),\varnothing\to xb}^{\Omega,\beta} }{RT}}
			\cdot\Y{i-1,k+1}{m-\Kron_{xb,s_is_k-m'}}{x,b}
			\cdot\Z{i+1,k-1}{m'}{x,b}
		&\text{If }S_i=k>i\\
    \displaystyle
			\sum_{m=0}^{M}
			\sum_{\substack{b\in B\\\Kron_{bx,s_ks_i\leq m}}}
			\sum_{m'=0}^{m-\Kron_{bx,s_ks_i}}
     	 e^{\frac{-E^{\Omega, \beta}_{(k,i),\varnothing\to bx}}{RT}}
			\cdot\Y{k-1,i+1}{m-\Kron_{bx,s_ks_i-m'}}{b,x}
			\cdot\Z{k+1,i-1}{m'}{b,x}
		&\text{If }S_i=k<i
	\end{array}\right.
$}\label{eq:combine}
\end{equation}
